|

### 2021-10-06 article/article.Rmd
update references
some copyediting

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of particular legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves the provision of public goods. We argue that changes in bargaining costs depend on whether additional legislators share the Executive's party affiliation: More opposition members reduce the Executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminish elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

@@

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider an strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies that is voted by the city councilors. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

@@

The mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that the selectorate, that is, the individuals who have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate ($N/2$). City councilors are motivated by policy concerns $p$ and transfers received from the mayor or the councilor recognized as the proposer \textbf{x}. We leave the definition of these transfers open, as this allows us to analyze how mayors combine transfers with other incentives to pressure city councilors. Exemples of transfers are granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even giving campaign contributions and bribes. The use of transfers to influence the councilors' choices generate a bargaining cost of $C_G$. Bargaining costs vary according to the reversal mechanism and the size of the legislature.

@@

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by $\delta$, and another city councilor is recognized to propose. The process repeats until a proposal is finally accepted.

@@

In equilibrium, the budget binds, and the first-order condition for an optimal public goods provision makes the marginal benefits to increase rents. In this case, it means that the marginal costs of providing the public goods is equal to the marginal benefits of reelection:

@@

\begin{proposition}\label{prop1}
The provision of public goods increases with legislature size when the bargaining costs decrease with legislature size.
\end{proposition}

@@

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then the provision of the public goods increases when the legislature size increase}. Now we study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

@@

The intuition for the results is that every additional councilor demands more resources from the mayor. Thus, more councilors represent higher bargaining costs. And if the bargaining costs increase with legislature size, then any increase in public goods provision associated with legislature size has come from a different mechanism. Below we show that partisanship is the key to understand how larger legislature sizes can increase public goods provision.

@@

Second, more seats could increase the competitiveness in the election, making politicians provide more services \citep{arvate2013electoral}. In a political system with around thirty parties and a sample average of almost ten parties in the municipalities we study, candidates per seat can increase exponentially. According to the Brazilian electoral law, each party can sponsor 1.5 times the number of seats as candidates. If a party form a pre-electoral coalition, this number increases to two times the number of seats \citep{coneglian2008lei}. An election in a small town, with ten councilors, may end up with 600 candidates: (30 parties) x (two times the number of seats) x (ten seats). Therefore, an alternative hypothesis to explain the increased service provision is that larger council size increase the competitiveness of the election.\footnote{Despite the common assumption that competitiveness improve welfare, \citet{boulding2014political} and \citet{gottlieb2019countervailing} can increase budgetary pressure, lower social policy, lead to bargaining inefficiencies.}

@@

The last set of variables quantify social welfare. Our focus is on the impact of legislature size on education and healthcare, which are the two most important public services the municipal government should provide to citizens. We measure access to education with the average enrollment in primary (K--4) municipal public schools. We assess education quality with the Education Development Index (\ 'Indice de Desenvolvimento da Educa\c{c}\~{a}o B\'asica--IDEB), which the Ministry of Education has issued biannually since 2007. The index is a weighted average of student scores in Portuguese and Mathematics (0--10), multiplied by the harmonic mean of public school promotion rates in a given year (0--100). To test the effect of the 2004 ruling on healthcare, we collect data on infant mortality per thousand children born and on postneonatal infant mortality, which is the death of infants who survived more than 28 days, but died before reaching one year of age. The source of all health care variables is the Brazilian Ministry of Health Data Center (DataSUS).

@@

To avoid this problem, in section A.5 of the Supplemental Materials, we show by simulation that adding controls, especially the variables responsible for the multiple threshold assignments (in our case, the population in 2003), improves the consistency and efficiency of the estimates. Relying on this fact, we add five controls to the estimates: population in 2003; GDP per capita; the number of seats in 2000; year; and a dummy for the northeast municipalities. The reason for the first variable is the multiple threshold assignments. We add GDP per capita because wealthier cities tend to have better public services. The number of seats in 2000 intends to control that some municipalities could experience a change provoked by the previous council size, confounding our estimates. The year and the Northeast dummies serve to improve efficiency. Adding controls improves efficiency in RDD and also enhances the consistency in the multiple thresholds RDD \citep{calonico2019regression}.

For the last assumption, there should be no changes in pre-treatment covariates. These pre-treatment covariates are variables measured before the primary outcomes. As the threshold rule selected by the Brazilian Electoral Court was unpredictable, it should not detect any variation before the thresholds were in place.

Table \ref{validitycheck} displays the results of the validity checks. We present the first-stage regressions at the top of the Table, with and without covariates, for both real and placebo cutoffs. At the bottom, we estimate the pre-treatment covariate balance. Furthermore, in the Supplementary Materials, we run placebo checks for all the models using placebo cutoffs within the real ones. The results are available in sections A.6 to A.12.

@@

We built two novel datasets to provide evidence for our mechanism. First, we ran a survey with former city councilors, asking how the mayor consolidate its coalition support within the city council. Second, we collected 346,553 laws approved in 63 municipalities within ten thousand inhabitants away from the city council size thresholds. We hand-coded 3,466 laws into four categories: public goods, oversight, education and health care-related legislation, and other legislative duties. After hand-coding these laws, we used a Supporting Vector Machine (SVM) algorithm to classify the remainder laws. From all of the algorithms we tested, SVM achieved the highest performance.

@@

We have established in Table \ref{tab:bargain} column (4) that the quantity of laws councilors approve remain constant. However, this tells us little about the legislation composition in municipalities with larger council sizes. In Figure \ref{LEGAPPR}, we show that the per capita number of laws related to local public goods provision increases in towns that receive the extra legislator, vis-à-vis baseline cities that did not get an additional councilor. One additional city councilor increased the per capita amount of laws on public goods by 15.0 percent. Also, the number of laws that focus on healthcare and education increased by 19.7 percent. Looking specifically into healthcare and education legislation, the proportion of laws discussing public goods increased by 15.7 percent, showing that policy proposals by both the mayor and the councilors shift toward local and citywide public goods.\footnote{Public goods here are not restricted to citywide goods. Any service that benefits citizens, regardless of whether they are targeted to citizens or not, is considered public goods. For example, building a health clinic in a particular district can virtually exclude the citizens that are not living in the neighborhood. However, we am still considering these as public goods, albeit targeted at a specific locality.}

@@

In terms of education, with the addition of a city councilor, enrollment in elementary schools increased by 2.58 children on average per school. This represents a change of around 0.20 standard deviations. Besides the influence of the mayoral support, enrollment can also directly impact the councilor's representation: Councilors can pressure the bureaucracy to accommodate a few extra students. With support from the mayor, they can help voters get access to schools, improving the voter's welfare. The education quality remains unchanged, but with more students in the classroom, it could potentially decrease. As expected, the placebo regressions are insignificant.

@@

\citet{Ames2001} argues that the Brazilian democracy is deadlocked because the Executive has its hands tied by a legislature that demand constantly provision of particularistic and private goods. Indeed, the most successful presidents after 1995 had to build multiparty coalitions by negotiating with parties \citep{limongi1998bases, raile2011executive}. Despite evidence that side payments cemented these coalitions, many social and economic policies have been implemented. For instance, education and health care notably improved during the Cardoso and Lula governments \citep{acemoglu2012nations}. Of the presidents that defied the coalition system, Rousseff was impeached in her second term. Bolsonaro can be considered one of the most ineffective presidents in Brazilian history.

@@

This paper opens two questions for future analysis. First, if the mechanism we suggest is indeed at work, more support for the Executive means better public goods. Therefore, one may ask what should be the city council's checking prerogatives, as the they significantly reduce the provision of public services. The evidence so far is inconclusive, and \citet{poulsen2018political} show that checks for the mayors may also improve service provision. It may be the relationship between rents and public goods is non-linear, making mayors revert to only rents past a certain threshold. Moreover, it could be that, at some point, legislative support is ineffective or even harmful to democracies, as it may trigger an intense form of elite capture. This question needs further investigation, and our results suggest that mayors with more support improve welfare. Second, it remains uncertain how other legislative features affect welfare. For example, city councils may vary in size, monetary compensation for councilors, committee structure, and internal power structure. These components may interact with each other, and understanding how size plays with other features would improve our knowledge about city councils size affect welfare. Both these questions are crucial for institution design and welfare in developing democracies.

>>>>

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of particular legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves public goods provision. We argue that changes in bargaining costs depend on whether additional legislators share the Executive's party affiliation: More opposition members reduce the Executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminishing elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

@@

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider a strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies on which city councilors vote. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

@@

The mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that the selectorate, that is, the individuals who have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate ($N/2$). City councilors are motivated by policy concerns $p$ and transfers received from the mayor or the councilor nominated as the proposer \textbf{x}. We leave the definition of these transfers open, as this allows us to analyze how mayors combine transfers with other incentives to pressure city councilors. Exemples of transfers are granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even giving campaign contributions and bribes. The use of transfers to influence the councilors' choices generate a bargaining cost of $C_G$. Bargaining costs vary according to the reversal mechanism and the size of the legislature.

@@

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by $\delta$, and another city councilor is nominated to propose. The process repeats until a proposal is finally accepted.

@@

In equilibrium, the budget binds, and the first-order condition for optimal public goods provision indicates that the marginal costs of providing public goods is equal to the marginal benefits of reelection:

@@

\begin{proposition}\label{prop1}
Public goods provision increases with legislature size when the bargaining costs decrease with legislature size.
\end{proposition}

@@

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then public goods provision increases when the legislature size increase}. Now we study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

@@

The intuition for the results is that every additional councilor demands more resources from the mayor. Thus, more councilors represent higher bargaining costs. And if the bargaining costs increase with legislature size, then any increase in public goods provision associated with legislature size has come from a different mechanism. Below we show that partisanship is the key to understand how this relationship.

@@

Second, more seats could increase the competitiveness in the election, making politicians provide more services \citep{gottlieb2019countervailing}. In a political system with around thirty parties and a sample average of almost ten parties in the municipalities we study, candidates per seat can increase exponentially. According to the Brazilian electoral law, each party can sponsor 1.5 times the number of seats as candidates. If a party forms a pre-electoral coalition, this number increases to two times the number of seats \citep{coneglian2008lei}. An election in a small town, with ten councilors, may end up with 600 candidates: (30 parties) x (two times the number of seats) x (ten seats). Therefore, an alternative hypothesis to explain the increased service provision is that larger council size increase the competitiveness of the election.\footnote{Despite the common assumption that competitiveness improve welfare, \citet{boulding2014political} and \citet{gottlieb2019countervailing} can increase budgetary pressure, lower social policy, lead to bargaining inefficiencies.}

@@

The last set of variables quantify social welfare. Our focus is on the impact of legislature size on education and health care, which are the two most important public services the municipal government should provide to citizens. We measure access to education with the average enrollment in primary (K--4) municipal public schools. We assess education quality with the Education Development Index (Índice de Desenvolvimento da Educação Básica--IDEB), which the Ministry of Education has issued biannually since 2007. The index is a weighted average of student scores in Portuguese and Mathematics (0--10), multiplied by the harmonic mean of public school promotion rates in a given year (0--100). To test the effect of the 2004 ruling on health care, we collect data on infant mortality per thousand children born and on postneonatal infant mortality, which is the death of infants who survived more than 28 days, but died before reaching one year of age. The source of all health care variables is the Brazilian Ministry of Health Data Center (DataSUS).

@@

To avoid this problem, in section A.5 of the Supplemental Materials, we show by simulation that adding controls, especially the variables responsible for the multiple threshold assignments (in our case, the population in 2003), improves the consistency and efficiency of the estimates. Relying on this fact, we add five controls to the estimates: population in 2003; GDP per capita; the number of seats in 2000; year; and a dummy for the northeast municipalities. The reason for the first variable is the multiple threshold assignments. We add GDP per capita because wealthier cities tend to have better public services. The number of seats in 2000 intends to control that some municipalities could experience a change provoked by the previous council size, confounding our estimates. The year and the Northeast dummies were added to improve efficiency. Adding controls improves efficiency in RDD and also enhances the consistency in the multiple thresholds RDD \citep{calonico2019regression}.

For the last assumption, there should be no changes in pre-treatment covariates. These pre-treatment covariates are measured before the primary outcomes. As the threshold rule selected by the Brazilian Electoral Court was unpredictable, it should not detect any variation before the thresholds were in place.

Table \ref{validitycheck} displays the results of the validity checks. We present the first-stage regressions at the top of the Table, with and without covariates, for both real and placebo cutoffs. At the bottom, we estimate the pre-treatment covariate balance. Furthermore, in the Supplementary Materials, we run placebo checks for all the models using placebo cutoffs within the real ones. The results are available in sections A.6 to A.12 of the Supplementary Materials.

@@

We built two novel datasets to provide evidence for our mechanism. First, we ran a survey with former city councilors, asking how mayors consolidate their coalition support within the city council. Second, we collected 346,553 laws approved in 63 municipalities within ten thousand inhabitants away from the city council size thresholds. We hand-coded 3,466 laws into four categories: public goods, oversight, education and health care-related legislation, and other legislative duties. After hand-coding these laws, we used a Supporting Vector Machine (SVM) algorithm to classify the remainder laws. From all of the algorithms we tested, SVM achieved the highest performance.

@@

We have established in Table \ref{tab:bargain} column (4) that the quantity of laws councilors approve remain constant. However, this tells us little about the legislation composition in municipalities with larger council sizes. In Figure \ref{LEGAPPR}, we show that the per capita number of laws related to local public goods provision increases in towns that receive the extra legislator, vis-à-vis baseline cities that did not get an additional councilor. One additional city councilor increased the per capita amount of laws on public goods by 15.0 percent. Also, the number of laws that focus on health care and education increased by 19.7 percent. Looking specifically into health care and education legislation, the proportion of laws discussing public goods increased by 15.7 percent, showing that policy proposals by both the mayor and the councilors shift toward local and citywide public goods.\footnote{Public goods here are not restricted to citywide goods. Any service that benefits citizens, regardless of whether they are targeted to citizens or not, is considered public goods. For example, building a health clinic in a particular district can virtually exclude the citizens that are not living in the neighborhood. However, we am still considering these as public goods, albeit targeted at a specific locality.}

@@

In terms of education, with the addition of a city councilor, enrollment in elementary schools increased by 2.58 children on average per school. This represents a change of around 0.20 standard deviations. Besides the influence of the mayoral support, enrollment can also directly impact the councilor's representation: Councilors can pressure the bureaucracy to accommodate a few extra students. With support from the mayor, they can help voters get access to schools, improving the voter's welfare. Education quality remains unchanged, but with more students in the classroom, it could potentially decrease. As expected, the placebo regressions are insignificant.

@@

\citet{Ames2001} argues that the Brazilian democracy is deadlocked because the Executive has its hands tied by a legislature that demands constantly provision of particularistic and private goods. Indeed, the most successful presidents after 1995 had to build multiparty coalitions by negotiating with parties \citep{limongi1998bases, raile2011executive}. Despite evidence that side payments cemented these coalitions, many social and economic policies have been implemented. For instance, education and health care notably improved during the Cardoso and Lula governments \citep{acemoglu2012nations}. Of the presidents that defied the coalition system, Rousseff was impeached in her second term. Bolsonaro can be considered one of the most ineffective presidents in Brazilian history.

@@

This paper opens two questions for future analysis. First, if the mechanism we suggest is indeed at work, more support for the Executive means better public goods. Therefore, one may ask what should be the city council's checking prerogatives, as they significantly reduce the provision of public services. The evidence so far is inconclusive, and \citet{poulsen2018political} show that checks for the mayors may also improve service provision. It may be the relationship between rents and public goods is non-linear, making mayors revert to only rents past a certain threshold. Moreover, it could be that, at some point, legislative support is ineffective or even harmful to democracies, as it may trigger an intense form of elite capture. This question needs further investigation, and our results suggest that mayors with more support improve welfare. Second, it remains uncertain how other legislative features affect welfare. For example, city councils may vary in size, monetary compensation for councilors, committee structure, and internal power structure. These components may interact with each other, and understanding how size plays with other features would improve our knowledge about city councils size affect welfare. Both these questions are crucial for institution design and welfare in developing democracies.
